Discovering causal factors with back-to-back regression

### 2019-05-22 paper/paper.tex
backwards theorem

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
\usepackage[ruled, linesnumbered]{algorithm2e}

\usepackage[hidelinks]{hyperref}
\usepackage[]{neurips_2019}
\DeclareMathOperator{\Tr}{Tr}

\SetKwComment{Comment}{$\triangleright$\ }{}

@@

\title{Measuring causal influence with\\ back-to-back regression}

@@

    \node[] (labY) at (\posY, 3){Observations};
    \node[] (labX) at (\posX, 3){Factors};
    \node[] (labE) at (\posE, 3.3){Cause};
    \node[] (labE) at (\posE, 3){selection};
    \node[] (labN) at (\posN, 3){Noise};
    \node[] (labF) at (\posF, 3.3){Cause-effect};
    \node[] (labF) at (\posF, 3){mapping};

    \node[] (labY) at (\posY, 2)   {\color{gray} MEG recordings};
    \node[] (labX) at (\posX, 2)   {\color{gray} stimulus};
    \node[] (labX) at (\posX, 2.3) {\color{gray} features};
    \node[] (labE) at (\posE, 2.3) {\color{gray} causal};
    \node[] (labE) at (\posE, 2)   {\color{gray} features};
    \node[] (labN) at (\posN, 2)   {\color{gray} breathing\ldots};
    \node[] (labF) at (\posN, 2.3) {\color{gray} bloodflow,};
    \node[] (labF) at (\posF, 2)   {\color{gray} MEG helmet};

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{Diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
    \end{tikzpicture}
    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and ii) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
    \label{fig:b2b}
\end{figure}

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.

Given the model in Equation~\eqref{eq:model}, the goal of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

%Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(EX+N)$ (f a function, E, X and N matrices). In particular, the addition of "external noise" M, as in $Y=F(EX+N)+M$ can be viewed as a small change in F and N.

%We consider both the linear and non linear case. In the first one, our model is $$Y = F(EX+N)$$, with  $Y\in\mathbb{R}^{n\times ny}$, $F\in\mathbb{R}^{ny\times nx}$, $X\in\mathbb{R}^{nx\times n}$, $N\in\mathbb{R}^{nx\times n}$, and E a binary diagonal square matrice of size nx.

% In the non linear model, $F(EX+N)$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma (F(EX+N))$$

%Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.

@@

Back-to-Back Regression (B2BR) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
%
This backward regression recovers an approximation of the causes of $Y$ in $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
%
In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

% If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E. That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.

% In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y). This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding). To guard against high condition numbers, we have used two methods: truncated SVD (ref) and ridge regression (ref).

% In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it. In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations. In ridge regression, a constant factor is added to all eigenvalues. The regularisation factor and truncation threshold are chosen via leave one out cross validation. In practice, ridge regression tends to provide better stability but introduces a bias in the estimates. Truncated SVD is used to provide better results in low noise cases.

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.

We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
%
The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
        $E \leftarrow E + \text{Diag}(H)$\;
    }
    \Return{$\hat{E} / m$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

\subsection{Theoretical guarantees}
As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with f an unknown function.
%
If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (dx, n) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
%
N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (dy, n) matrix of measured effects.

The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).

For the first step, replacing Y, and since N is centred, we have
\begin{equation}
\begin{aligned}
\arg \min_G \left \| X-GY \right \|^2 &= \arg \min_G \left \| X - GF(EX+N)\right\|^2 \\
&{}= \arg \min_G \left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2
\end{aligned}
\end{equation}

As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-GFEX\right\| ^2$, which is $X(FEX)^\dagger$ ($M^\dagger$ denoting the pseudo inverse of M).

Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $FEX$ be of rank k (which means F and X have full rank on the space spanned by the active features of E). Let K be the $\mathbb{R}^{nx\times k}$ diagonal matrix obtained by deleting the rightmost nx-k columns of E. Then, $E=KK'$, and $FEX=(FK)(K'X)$, with $FK$  and $K'X$ of full rank k.

@@

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
X(FEX)^\dagger &=X((FK)(K'X))^\dagger \\
&=X(K'X)'((K'X)(K'X)')^{-1}(FK)^\dagger \\
&=(XX')K(K'XX'K)^{-1}(FK)^\dagger 
\end{aligned}
\end{equation}

@@

As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
% \subsection{Asymptotic behaviour - non linear case}

\subsection{Use cases}

If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
%
This recovers from $\hat H$ a binary diagonal $\hat E$.
%
To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.

\section{Experiments}
\subsection{Synthetic data}

\input{synthetic.tex}

\subsection{Magnetoencephalography data}

\section{Related works}
\subsection{back-to-back regression as a special case of canonical component analysis?}
Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
%
In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

@@

% \section{Acknowlegements}
% Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.

@@

%\section{Appendices}
%
%This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data

>>>>

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage[]{neurips_2019}
\DeclareMathOperator{\Tr}{Tr}

@@

\title{Discovering causal factors with back-to-back regression}

@@

    \node[] (labY) at (\posY, 2){Observations};
    \node[] (labX) at (\posX, 2){Factors};
    \node[] (labE) at (\posE, 2.3){Cause};
    \node[] (labE) at (\posE, 2){selection};
    \node[] (labN) at (\posN, 2){Noise};
    \node[] (labF) at (\posF, 2.3){Cause-effect};
    \node[] (labF) at (\posF, 2){mapping};

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
    \end{tikzpicture}
    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$. }
    \label{fig:}
\end{figure}

\subsection{Formalism}
Via an apparatus F, we are measuring a multidimensional signal Y, which is a function of possible causes X and other non causal features N. Not all features of X exert causal influence on Y. Let $XE$ (E a matrix of zero and one) the subset of causal factors. Our model can be formalised as $Y=f(XE,N)$.

Without loss of generality, we may assume N to be centred(adding, when necessary, an intercept to the causal features), and redefine features and noise so that the model be written $Y=f(XE+N)$ (f a function, E, X and N matrices). In particular, the addition of "external noise" M, as in $Y=(XE+N)F+M$ can be viewed as a small change in F and N.

We consider both the linear and non linear case. In the first one, our model is $$Y = (XE+N)F$$with  $Y\in\mathbb{R}^{n\times n_{y}}$, $F\in\mathbb{R}^{n_{x}\times n_{y}}$, $X\in\mathbb{R}^{n\times n_{x}}$, $N\in\mathbb{R}^{n\times n_{x}}$, and E a binary diagonal square matrix of size $n_{x}$.

In the non linear model, $(XE+N)F$ is defined as before, but Y is the result of a non linear transformation of it $$Y=\sigma ((XE+N)F)$$

Under this formalism, we are given samples of Y and X, and are tasked to recover E and use it to predict Y from X.

@@

Back-to-back regression consists of two steps.
%
First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=Y\hat G$ is the least square predictor of X.
%
This recovers an approximation of the causes of Y in X.
%
Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
%
In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}

If, in the above formalism, F is invertible with respect to the causal features of X (ie the problem has a solution), and the relation is linear or close to linear, the diagonal of $\hat H$ will be a scaled approximation of the diagonal of E. That is to say, back to back regression will recover the causal features of the process considered, even in the presence of heavy noise N, and without needing to estimate the transformation F.

In practical cases, the covariance matrices of Y and X will often be badly conditioned (because features of X are related, or measurement produces correlated features of Y). This makes the inversion of $X'X$ and $Y'Y$ sensitive to noise (either from N, from sampling or from computation rounding). To guard against high condition numbers, we have used two methods: truncated SVD (ref) and ridge regression (ref).

In both, we compute the singular value decompositions of the covariance matrix, transform its diagonal part, and invert it. In truncated SVD, the lowest eigenvalues are set to zero, and the corresponding vectors eliminated from future calculations. In ridge regression, a constant factor is added to all eigenvalues. The regularisation factor and truncation threshold are chosen via leave one out cross validation. In practice, ridge regression tends to provide better stability but introduces a bias in the estimates. Truncated SVD is used to provide better results in low noise cases.

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data are detected in the first step and learnt in the second. To avoid this, we split our sample into two halves, and each regression is performed over one, effectively decorrelating the two steps.
%
To compensate for the reduction in sample size caused by the split, this process is repeated over many splits, and their results are averaged.
This bagging compensates for the reduction in sample size caused by splitting.

As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.

If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we select the split that maximises the Sonquist and Morgan criterion $$K = \frac{p (n-p)}{n} \frac{(s-r)^2}{ V}$$.
%
This recovers from $\hat H$ a binary diagonal $\hat E$.
%
To use back-to-back regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $X \hat E $.

Summarizing, back-to-back regression is performed as follows:
\begin{enumerate}
\item Randomly split the data sample.
\item From the first half sample, using cross-validated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
\item From the second half sample, calculate $\hat X = Y \hat G$, and find $\hat H$, the regularised least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
\item Repeat steps 1 to 3 and average the results as $\hat H$.
\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $X \hat E$.
%

\end{enumerate}

\subsection{Asymptotic behaviour - linear case}
As discussed above, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(XE+N)$, with f an unknown function.
%
If F is linear we can rewrite it as $Y = (XE + N)F$, with $X\in\mathbb{R}^{n\times n_{x}}$ a matrix of possible causes, E a  binary diagonal $n_{x}$ square matrix. selecting active features of X.
%
$N\in\mathbb{R}^{n\times n_{x}}$, a homoscedastic noise matrix, $F\in\mathbb{R}^{n_{x}\times n_{y}}$ an unknown transformation corresponding to the measuring apparatus, and $Y\in\mathbb{R}^{n\times n_{y}}$ a matrix of measured effects.

The two steps of back-to-back regression amounts to finding $\hat G=\arg \min_G \left \| X-YG \right \|^2$ and $\hat H =\arg \min_H \left \| \hat YG - XH \right \|^2$ (all matrix norms henceforth are Frobenius norms).

For the first step, replacing Y, and since N is centred, we have
\begin{equation}
\begin{aligned}
\arg \min_G \left \| X-YG \right \|^2 &= \arg \min_G \left \| X - (XE+N)FG\right\|^2 \\
&{}= \arg \min_G \left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2
\end{aligned}
\end{equation}

As N tends to zero, we are searching for the minimal norm solution of $\min_G \left \| X-XEFG\right\| ^2$, which is $(XEF)^\dagger X$ ($M^\dagger$ denoting the pseudo inverse of M).

Let k be the number of active features of E, suppose for notational convenience that they are the first k features of X, and let $XEF$ have rank k (which means F and X have full rank on the space spanned by the active features of E). Let K be the $\mathbb{R}^{k\times N_{x}}$ diagonal matrix obtained by deleting the bottom $n_{x}-k$ lines of E. Then, $E=K'K$, and $XEF=(XK')(KF)$, with $KF$  and $XK'$ of full rank k.

@@

Replacing in the expression of the minimal norm solution yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK')(KF))^\dagger X\\
&=(FK)^\dagger (XK')'((XK')(XK')')^{-1} X\\
&=(FK)^\dagger KX' (K'XX'K)^{-1} X
\end{aligned}
\end{equation}

@@

\subsection{Asymptotic behaviour - non linear case}

\section{Related works}
\subsection{Back-to-back regression as a special case of canonical component analysis}
Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
%
In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

\section{Experiments - simulated data}
\input{synthetic.tex}

\section{Experiments - real case}
Jean Remi and Maxime territory

@@

\section{Acknowlegements}
Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.

@@

\section{Appendices}

This should hold part of the tests, explanations on simulation, detailed results and stuff on meg data
